### 2023-01-05 manuscript.Rmd
update manuscript for initial submission

---
output: word_document
bibliography: doc/references.bib
csl: doc/nature.csl
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for Alzheimer's Disease (AD).[@jansen2019; @wightman2021; @bellenguez2022] Most AD GWAS have focused on clinical AD; however, clinical AD does not always correspond with AD neuropathology and vice-versa.[@nelson2012] We hypothesized that investigating the genetic architecture of multiple neuropathological endophenotypes (NPE) would:

1.  identify novel genetic risk loci for individual NPE
2.  identify common sources of genetic risk (pleiotropy) between NPE
3.  provide evidence that the effects of clinical AD genetic risk loci may be mediated by specific NPE.

@@

We used genotype and NPE data from four aged autopsy cohorts: National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Memory and Aging Project (ROSMAP), Adult Changes in Thought (ACT), and the AD Neuroimaging Initiative (ADNI).

## Quality Control

Quality control and inclusion/exclusion criteria closely follow that used in our previous brain arteriolosclerosis GWAS,[@shade2022] with the exception that all cohorts have undergone imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel.[@taliun2021]

1.  Retrieve participants labeled as non-Hispanic white or similar (European, etc.) if available.
2.  Perform imputation on TOPMed Imputation Server (ROSMAP and ADNI had already been imputed using same reference panel by collaborators, so this step was skipped for those cohorts).
3.  Covert VCF files to PLINK file set, keeping variants will minor allele frequencies (MAF) of 0.1% or greater.
4.  Merge sub-cohorts of same study (e.g. NACC cohorts 1-12 and ACT 1-3).
5.  Remove duplicate participants in each study.
6.  Remove participants without autopsy data available.
7.  Iteratively remove variants and participants with excess missingness until no variant is missing in more than 5% of participants and no participant is missing more than 5% of variants.
8.  Remove participants with unusually high/low heterozygosity (+/- 3 SD cutoff).
9.  Merge with 1000 Genomes Phase 3 cohorts[@the1000genomesprojectconsortium2012] and remove participants without primarily European ancestry as determined by principal component analysis.
10. Remove participants with rare neurological diseases that substantially raise the likelihood that present NPE could be due to unusual pathological processes (NACC only, as the other cohorts did not have this information available).
11. Remove duplicate participants between studies, preferentially keeping the observation in the larger cohort (*i.e.* NACC \> ROSMAP \> ACT \> ADNI in preference).
12. Merge cohorts.
13. Harmonize NPE variables to analyze.
14. Create list of related ($\ge$ 2nd-degree relatives) individuals using KING[@manichaikul2010] to remove for each NPE variable, randomly keeping one participant in each related cluster.
15. Create genetic principal components using the PC-AiR method in the GENESIS package in R 4.1.[@conomos2015]

We note that there are substantial differences in the way that NPE are collected in different cohorts,(cite) and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes and minimizing computational load for analysis. Thus, some ordinal variables were collapsed to binary variables, or synthetic NPE may have been created using categorical variables from one cohort and continuous variables from another. All code used for harmonization will be made available upon publication, and a brief overview of the harmonization strategy for each NPE will be included in the Supplemental Materials.

@@

Single-variant logistic regression analyses were performed in PLINK 2.0.[@chang2015; @purcell] Covariates included age at death, sex, cohort, and the first 10 genetic principle components. Continuous covariates were standardized to a mean of 0 and a variance of 1. An additive mode of inheritance was assumed in all analyses. Variants achieving a p-value of $< 5\times 10^{-8}$ were considered "genome-wide" significant. To identify independent genetic risk loci, variants were clumped using PLINK 1.9[@chang] using a linkage-disequilibrium threshold of $r^2 \le 0.05$.

Ordinal variables were analyzed using proportional odds logistic mixed-effects models using the POLMM R package.[@bi2021] Dense kinship matrices were used to account for relatedness.

### Conditional and mediation analyses

For genetic loci associated with multiple NPE, we considered mediation hypotheses to test whether the association of the lead variant with one NPE was mediated by another NPE. Regression analyses were run in R using the `glm` or `polr` functions and mediation analyses were performed using the mediation[@tingley2014] R package with standard errors estimated using boot strap re-sampling with 1000 simulations.

@@

All variants in each logistic GWAS with p-values $< 0.01$ were checked for significant quantitative trait locus (QTL) activity in the Genotype-Tissue Expression Project (GTEx) v8 European ancestry data set in 48 tissues. Colocalization analysis using the coloc R package (cite) was then performed on GWAS loci with significant QTLs. A prior probability of colocalization of $1\times 10^{-5}$ was used, and a posterior probability of colocalization (PPH4) \> 80% was used as a threshold for evidence of colocalization.

>>>>

---
output: word_document
bibliography: doc/references.bib
csl: doc/springer_basic.csl
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(data.table)
library(flextable)
```

Proposed journal: ***Molecular Neurodegeneration***

**Title**

Genome-wide association study of multiple neuropathology endophenotypes identifies novel risk loci and provides insights into known Alzheimer's risk loci

**Authors**

Lincoln M.P. Shade^1^

Yuriko Katsumata^1,2^

Steven A. Claas^1^

Mark T. W. Ebbert^2,3,4^

Erin L. Abner^2,5^

Timothy J. Hohman^6^

Shubhabrata Mukherjee^7^

Kwagsik Nho^8^

Andrew J. Saykin^8^

David A. Bennett^9^

Julie A. Schneider^10^

Peter T. Nelson^2,11^

David W. Fardo^1,2^

for the Alzheimer's Disease Neuroimaging Initiative^\*^

**Author Affiliations**

^1^Department of Biostatistics, College of Public Health, University of Kentucky, Lexington, KY

^2^Sanders-Brown Center on Aging and Alzheimer's Disease Research Center, University of Kentucky, Lexington, KY

^3^Division of Biomedical Informatics, Department of Internal Medicine, University of Kentucky, Lexington, KY

^4^Department of Neuroscience, University of Kentucky College of Medicine, Lexington, KY

^5^Department of Epidemiology and Environmental Health, College of Public Health, University of Kentucky, Lexington, KY

^6^Vanderbilt Memory & Alzheimer's Center, Department of Neurology, Vanderbilt University Medical Center, Nashville, Tennessee

^7^Department of Medicine, University of Washington, Seattle, WA

^8^Department of Radiology & Imaging Sciences, Indiana University School of Medicine, Indianapolis, IN

^9^Rush Alzheimer's Disease Center and Department of Neurological Sciences, Rush Medical College, Chicago, IL

^10^Rush Alzheimer's Disease Center and Departments of Neurology and Pathology, Rush University Medical Center, Chicago, IL

^11^Department of Pathology and Laboratory Medicine, University of Kentucky, Lexington, KY

^\*^Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at: <http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf>

**Corresponding Author**

David W. Fardo

Healthy Kentucky Research Building

760 Press Ave, Room 372

Lexington, KY 40536-0679

[david.fardo\@uky.edu](mailto:david.fardo@uky.edu)

859-218-2070

**Abstract**

***Background***

Alzheimer's disease is highly heritable and exhibits neuropathological hallmarks of neurofibrillary tau tangles and neuritic amyloid plaques. However, the reality of the relationship between clinically diagnosed Alzheimer's disease and neuropathology is more complex. Upon autopsy, many Alzheimer's disease patients have multiple comorbid neuropathologies other than plaques and tangles which may have independent or pleiotropic genomic risk factors. Autopsy data combined with genome-wide association study (GWAS) provides the opportunity to study the genetic risk factors of multiple neuropathologies that contribute to dementia.

***Methods***

We studied the genome-wide risk factors of eleven Alzheimer's disease-related neuropathology endophenotypes. We used four sources of neuropathological data: National Alzheimer's Coordinating Center, Religious Orders Study and Rush Memory and Aging Project (ROSMAP), Adult Changes in Thought study, and Alzheimer's Disease Neuroimaging Initiative. We used generalized linear mixed models to identify risk loci, followed by Bayesian colocalization analyses to identify potential molecular functions of risk loci. We then analyzed associations between four CpG methylation sites, identified through colocalization analysis, with cerebral amyloid angiopathy (CAA) pathology in ROSMAP.

***Results***

After adjusting for *APOE* genotype, we identified a locus near *APOC2* (lead variant rs4803778) associated with CAA that colocalizes with DNA methylation at four nearby CpG sites in the cerebral cortex. Two of these sites, cg0955818 ($p = 0.004$) and cg13119609 ($p = 0.0007$), were themselves significantly associated with CAA in ROSMAP. We identified two other novel neuropathology risk loci: one *PIK3R5* locus (lead variant rs72807981) with neurofibrillary pathology, and one *COL4A1* locus (lead variant rs2000660) with cerebral atherosclerosis. We also confirmed associations between known Alzheimer's genes and multiple neuropathology endophenotypes, including *APOE* (neurofibrillary tangles, neuritic plaques, diffuse plaques, cerebral amyloid angiopathy, and TDP-43 pathology); *BIN1* (neurofibrillary tangles and neuritic plaques); and *TMEM106B* (TDP-43 pathology and hippocampal sclerosis).

***Conclusions***

Associations between the Chromosome 19 CAA risk locus, CpG methylation, and CAA pathology are consistent with the effect of this locus on disease being mediated by DNA methylation. Furthermore, the same two CpG sites are associated with dementia in an independent cohort, indicating that genomic signals other than *APOE* $\epsilon$ alleles in the *APOE* region may independently affect dementia risk. The atherosclerosis risk variant rs2000660 is in strong linkage disequilibrium with a synonymous coding variant (rs650724) of *COL4A1*, providing a candidate functional variant. *BIN1* is associated with neurofibrillary tangles and neuritic plaques but not with diffuse amyloid pathology. *TMEM106B* is associated with hippocampal sclerosis and TDP-43 pathology but not the canonical Alzheimer's disease pathologies. Our findings provide novel insights into neuropathology risk factors and parse clinical Alzheimer's genetic risk into specific key neuropathological domains.

# Key words

GWAS, Alzheimer's, neuropathology, atherosclerosis, neurofibrillary tangles, cerebral amyloid angiopathy, TDP-43

# Background

Aged individuals often accumulate multiple brain pathologies that contribute individually and synergistically to cognitive decline and dementia. Alzheimer's disease (AD) is the most commonly diagnosed form of dementia and poses an enormous burden on human health and well-being. In the United States alone, over six million individuals are living with AD, and the disease imposes financial costs of treatment and care of over \$300 billion annually [@2021alz2021].

The most common form of AD, late-onset AD (LOAD), presents in individuals aged 65 years and older and is highly heritable, with twin and family studies estimating its heritability at \~60% [@ridge2013; @gatz1997; @gatz2006]. Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for LOAD that are involved biological processes including amyloid precursor protein processing, immune response, and extracellular matrix regulation [@jansen2019; @wightman2021; @bellenguez2022]. Most dementia GWAS to date have focused on clinically diagnosed LOAD or proxy outcomes based on family history of dementia [@jansen2019; @wightman2021; @bellenguez2022; @kunkle2019]; however, late-onset dementia diagnosed as LOAD is increasingly recognized as a heterogeneous clinical syndrome that can reflect multiple underlying independent or co-occurring pathological processes that may have independent or pleiotropic genetic risk factors [@farfel2016; @katsumata2022; @karanth2020].

While neuritic plaques and neurofibrillary tangles (NFT) are present in the brains of the majority of people diagnosed with clinical AD, many people either do not have substantial AD neuropathology or have both AD and comorbid non-AD pathologies [@nelson2009; @karanth2020]. For example, TDP-43 pathology contributes to an amnestic syndrome, limbic-predominant age-related TDP-43 encephalopathy (LATE), that presents similarly to LOAD [@nelson2019]. However, LATE and LOAD exhibit differing patterns of pathogenesis in aged autopsy cohorts. LOAD pathology peaks in prevalence in individuals who die at age 95 years and decreases thereafter [@farfel2019]. In contrast, prevalence of LATE neuropathologic change (LATE-NC) increases monotonically with age of death, with TDP-43 pathology often first appearing in the limbic region or entorhinal cortex and progressing to the neocortex [@farfel2019; @nelson2019]. Hippocampal sclerosis (HS) of aging, a pathology characterized by uni- or bilateral neuronal death, gliosis, and atrophy of the hippocampus beyond normal ranges based on levels of AD pathology, commonly co-occurs with LATE-NC and is associated with severe cognitive impairment [@brenowitz2014; @nelson2019].

In addition to proteinopathies, non-stroke vascular pathologies are common in elderly autopsied research participants and contribute to cognitive decline and dementia [@skrobot2016]. Cerebral amyloid angiopathy (CAA) is one such vascular pathology characterized by amyloid-beta deposition in cerebral blood vessels [@weber2018]. CAA often co-occurs with LOAD, but can independently contribute to cerebral injury by causing hemorrhages in the brain parenchyma [@weber2018; @skrobot2016]. Infarcts of both small but grossly visible lacunar arteries and microscopically examined vessels (the latter referred to as microinfarcts) are also common in aged individuals and contribute to cognitive decline [@smith2012; @skrobot2016]. Cerebral atherosclerosis and sclerosis of small cerebral blood vessels, called brain arteriolosclerosis, predispose individuals to infarcts and hippocampal sclerosis [@arvanitakis2017; @neltner2014]; moreover, they contribute to cognitive decline even after adjusting for the presence of related pathologies [@arvanitakis2016; @ighodaro2017]. Collectively, these factors reveal an increasingly complex and synergistic web of pathologies contributing to cognitive impairment and dementia, and demonstrate the need for continued study into genomic risk factors of dementia beyond those of clinically diagnosed LOAD.

The investigation of genetic risk of neuropathologies is an attractive approach for uncovering biological pathways involved in the development of LOAD and related dementias. Neuropathologies are commonly operationalized in aged autopsy cohorts as binary or semi-quantitative (or ordinal) neuropathology endophenotypes (NPE). Prior GWAS of neuropathology endophenotypes have confirmed known LOAD risk loci and have identified potential neuropathology risk loci [@beecham2014; @shade2022; @vattathil2021; @farrell2022]. However, continued recruitment in ongoing autopsy cohorts and methodological development for GWAS and downstream functional analysis provides additional opportunity for in-depth investigation into multiple neuropathologies studied across multiple cohorts.

In this study, we perform GWAS on eleven neuropathology endophenotypes collected in four high-quality studies with autopsy and genotype data available. Endophenotypes studied include AD-associated pathologies; LATE-NC; hippocampal sclerosis; vascular NPE including cerebral amyloid angiopathy, gross infarcts, microinfarcts, atherosclerosis, and arteriolosclerosis; and Lewy body pathology. We also perform downstream functional analyses to explore potential biological functional mechanisms of identified risk loci and provide insight into the shared genomic risk of neuropathologies.

@@

## Participants

```{r}
np <- readRDS("data/mega/mega_np.Rds")
nadrc <- np[, length(table(NACCADC))]
```

We used genotype and neuropathology data from four high-quality data sources: the National Alzheimer's Coordinating Center (NACC), the Religious Orders Study and the Rush Memory and Aging Project (ROSMAP), the Adult Changes in Thought (ACT) study, and the AD Neuroimaging Initiative (ADNI). Some participants in ROSMAP and ADNI also had neuropathology and genotype data available in NACC. In these cases, records in the NACC were preferentially kept in order to maximize sample size of the initial analysis using only NACC participants (see Statistical Analyses subsection).

All study participants were deceased and the resulting data de-identified for University of Kentucky investigators, and we exclusively used archival samples. Therefore, our study does not fall under the definition of "Human Subjects Research" according to the University of Kentucky Institutional Review Board because of NIH Exemption #4 -- "involving the collection/study of data or specimens if publicly available, or/or recorded such that subjects cannot be identified."

The present study used NACC data from `r nadrc` National Institute on Aging-funded Alzheimer's Disease Research Centers (ADRCs). Individual ADRCs use different recruitment strategies and perform autopsies on-site, but neuropathology data at each ADRC are collected using a standard form (<https://files.alz.washington.edu/documentation/np11-form.pdf>) and submitted to NACC where they are aggregated and anonymized. The NACC Neuropathology (NP) data set based on the first version of this form was originally implemented in 2001 [@besser2018], and this analysis uses data from then through the December 2021 freeze. Participants were excluded if they did not have autopsy data available or if they were noted in the NP data set to have at least one of 19 conditions that could potentially bias results. These conditions include brain tumors, severe head trauma, and fronto-temporal lobar degeneration. (See **Supplementary Table S1** for full list of variables used for exclusion criteria.)

ROSMAP has been previously described and consists of harmonized data from two longitudinal cohort studies: The Religious Orders Study (ROS) and the Rush Memory and Aging Project (MAP) [@bennett2018]. ROS and MAP were both approved by an Institutional Review Board of Rush University Medical Center. All participants signed an Anatomic Gift Act, as well as informed and repository consents. ROS began in 1994 and has recruited over 1500 Catholic priests, nuns, and brothers across the United States. MAP started in 1997 and has enrolled more than 2300 community members in the greater Chicago area of northeastern Illinois. The ROSMAP NP data used in this study were received from Rush University Medical Center in January 2020.

The ACT study began in 1994 and recruited residents in the greater Seattle area aged 65 years and older without dementia at time of enrollment [@kukull2002]. The study has expanded to include three cohorts with continued enrollment using the original enrollment criteria and has a current total of 4,960 participants across all three cohorts. The ACT NP data used in this study were obtained from Kaiser Permanente in May 2021.

The ADNI (<https://adni.loni.usc.edu>) was launched in 2003 as a public-private research partnership, led by Principal Investigator Michael W. Weiner, MD. The primary goal of ADNI has been to test whether serial magnetic resonance imaging, positron emission tomography (PET), other biological markers, and clinical and neuropsychological assessment can be combined to measure the progression of mild cognitive impairment and early AD. A subset of ADNI participants undergo autopsy and receive neuropathological phenotyping. The ADNI NP data used in this study was downloaded from the ADNI website in October 2021.

## Genotype data and quality control

Genotype data for all cohorts underwent imputation using the Trans-'Omics for Precision Medicine (TOPMed) Imputation Server and the TOPMed reference panel [@taliun2021]. NACC and ACT raw genotype data were obtained from the September 2020 freeze of the Alzheimer's Disease Genetics Consortium (ADGC) and subsequently imputed, while pre-imputed ROSMAP and ADNI genotype data were received from collaborators in the Hohman lab at Vanderbilt University in March 2021 and December 2021, respectively.

Quality control on genotype data was performed separately for each data source. Quality control and inclusion/exclusion criteria closely followed that used in our previous brain arteriolosclerosis GWAS [@shade2022]. Briefly, we included only participants labeled as non-Hispanic white or similar (European, etc.) if available. After imputation, we first identified and subsequently removed duplicate samples using the KING software "\--duplicate" option [@manichaikul2010]. We then removed participants without autopsy data available by merging genotype sample identification data with neuropathology data sets and retaining only samples present in both data sets. We then iteratively removed genetic variants and participants until no variants were missing in more than 5% of participants and no participants were missing more than 5% of variants (however, average genotype coverage was 99.7%, and no variants or participants were actually removed during this process). We then excluded participants with unusually high or low ($\pm$ 3 standard deviations from mean) genetic heterogeneity, as measured by the PLINK 1.9 "\--het" flag [@chang]. Finally, we merged participants with the 1000 Genomes Phase 3 (1000 Genomes) cohorts [@the1000genomesprojectconsortium2012] and performed principal components analysis (PCA) on a subset of independent variants (measured by pairwise $r^2 < 0.2$). We excluded participants with substantial non-European ancestry, as determined by distance from the 1000 Genomes EUR superpopulation centroid using the first two principal components (PCs).

## Defining and harmonizing neuropathology endophenotypes for analysis

In total, we combined or harmonized 11 neuropathology endophenotypes for analysis across the four studies. We note that there are differences in the way that some neuropathological data were collected across studies, and our strategy for harmonizing was informed by practical considerations for maximizing available samples sizes given the available endophenotypes. Thus, several synthetic NPE were created by merging existing NPE within a cohort or by harmonizing categorical variables from one cohort and continuous variables from another. A detailed listing of variables harmonized across cohorts to construct neuropathology endophenotypes for analysis is available in **Supplementary Table S3**. We briefly describe our harmonization methods below.

Arteriolosclerosis, Braak NFT stage, CAA, atherosclerosis, Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques, microinfarcts, and gross infarcts had variables in each cohort with directly comparable coding definitions and were straightforwardly renamed and combined with minimal recoding. Analyzed arteriolosclerosis, CAA, atherosclerosis, and CERAD score variables each had four stages with the following labels: 0 ("none"), 1 ("mild"), 2 ("moderate"), and 3 ("severe"). Microinfarcts and gross infarcts were labeled either 0 ("absent") or 1 ("present"). The Braak NFT Stage variable analyzed followed the staging criteria previously described in the literature and had seven levels, ranging from 0 (absent NFT) to six (diffuse NFT throughout cortex and large loss of neurons) [@braak1991].

LATE-NC was recorded differently in several of the data sets and was harmonized to a four-level outcome variable following the simplified staging of TDP-43 pathology outlined in Figure 3B of the 2019 LATE working group report [@nelson2019]. The following levels were used in the analyzed LATE-NC variable: 0, indicating lack of recorded TDP-43 proteinopathy; 1, indicating TDP-43 deposits in the amygdala only; 2, indicating deposits in the hippocampus or entorhinal cortex; and 3, indicating deposits in the neocortex. TDP-43 data was not available in ACT. In ROSMAP, TDP-43 pathology is recorded as a single variable following the same staging detailed above. In NACC and ADNI, the presence of TDP-43 pathology in each region of the brain is recorded as a separate binary indicator variable. To collapse TDP-43 pathology to a single ordinal variable in these data sets, we assigned a value based on the presence of the "highest" region where TDP-43 was present (*e.g.* a participant with TDP-43 pathology in both the hippocampus and the neocortex would be assigned a value of 3). Participants were labeled as 0 if they met two conditions: (1) they had recorded TDP-43 data available for at least one of the brain regions used for staging and (2) TDP-43 pathology was noted as absent in all the regions for which they had data available.

Diffuse amyloid plaque pathology was recorded as a four-stage Thal phase of amyloid deposition in NACC, ACT, and ADNI with the following levels: 0 ("none"), 1 ("mild"), 2 ("moderate"), and 3 ("severe"). In ROSMAP, diffuse plaques were examined and quantified in five regions (midfrontal cortex, entorhinal cortex, inferior parietal cortex, and hippocampus), then scaled by each region's standard deviation and averaged. To discretize this continuous variable in ROSMAP participants to a four-level variable, as recorded in the other data sets, we assigned participants a value of 0 ("none") if their averaged score was equal to 0, 1 ("mild") if their score was higher than 0 but $\le$ 0.5, 2 ("moderate") if their score was between 0.5 and one, and 3 ("severe") if their score was above 1. These labels roughly corresponded to score quartiles in ROSMAP.

Hippocampal sclerosis is recorded as a binary indicator of the presence or absence of pathology in the NACC NP data set form version 1-9, ROSMAP, and ACT. In versions 10 and 11 of the NACC NP form and in ADNI, HS pathology is recorded as being absent, unilateral, or bilateral. To harmonize, we dichotomized HS pathology as being present if either unilateral or bilateral pathology was indicated.

The Lewy body pathology variable we analyzed had four levels: 0, indicating absent Lewy body pathology in all regions examined or limited to the olfactory bulb; 1, indicating Lewy body pathology limited to the brainstem, including the substantia nigra; 2, indicating Lewy body pathology involving the limbic system or amygdala; and 3, indicating Lewy body involvement in the neocortex. In NACC and ROSMAP, Lewy body pathology was graded in ordinal variables with levels corresponding to the levels in the final harmonized outcome variable analyzed. In ADNI, a single variable was also used to record Lewy body pathology but included separate levels for (1) no pathology and olfactory-predominant and (2) limbic-predominant and amygdala-predominant pathology. These stages were collapsed into levels 0 and 2, respectively, for harmonization. In ACT, separate binary indicator variables were used to indicate presence of Lewy body pathology in each brain region checked. To harmonize, we created a new variable that coded Lewy body pathology stage according to the "highest" stage present in an individual (*e.g.* if pathology was present in both the amygdala and the neocortex, we assigned a value of 3).

@@

We analyzed ordinal endophenotypes using proportional odds logistic mixed-effects models implemented in the POLMM R package [@bi2021]. We analyzed binary variables similarly with logistic mixed-effects models implemented in the SAIGE R package [@zhou2018]. Covariates included age at death, sex, cohort, and the first 10 genetic PCs created using the PCA in Related Samples (PC-AiR) method in the GENESIS R package [@conomos2015]. We included a dense genetic relationship matrix (GRM) as a random effect to account for relatedness between participants. An additive mode of inheritance was assumed in all analyses.

Analyses proceeded in two stages. In stage one, GRM were constructed using a pruned set of independent variants, defined as having a pairwise $r^2 < 0.2$ within moving windows of 15 kilobase pairs (kbp). Null models including fixed covariates and the GRM were then fitted using either the POLMM or SAIGE R packages. In stage 2, score tests were performed on each variant with a saddle-point approximation used to calculate p-values. We considered all variants with a p-value of $p < 5\times 10^{-8}$ to be genome-wide significant. To identify independent risk loci, we clumped results using the "\--clump" flag in PLINK 1.9 with the pairwise linkage-disequilibrium (LD) threshold set to $r^2 \le 0.05$ [@chang; @chang2015].

We performed analyses both in individual cohorts separately and in a pooled data set with all participants. Single-variant analyses were first performed using NACC data. For all lead variants with a p-values $< 1 \times 10^{-6}$, we attempted to validate the association in ROSMAP, ACT, and ADNI separately. Participants were then pooled into a single data set and genome-wide association analyses were performed again using all participants. As noted in the section on harmonization above, LATE-NC data was not available for ACT participants, so validation of LATE-NC-associated variants was not attempted in ACT.

### Re-analysis of the *APOE* region

The region surrounding the *APOE* gene on Chromosome 19 is consistently the single strongest genetic risk factor for LOAD in GWAS. Three common forms of the *APOE* gene -- $\epsilon 2$, $\epsilon 3$, and $\epsilon 4$ -- are present in our study populations, and the $\epsilon 2$ and $\epsilon 4$ alleles are associated with lower and higher risk of LOAD, respectively, relative to the $\epsilon 3$ allele. We therefore expected that variants in the *APOE* region, defined as the region within 200 kbp from the start and end transcription sites of *APOE*, would be associated with multiple NPE in our study. Moreover, we hypothesized that genetic variants in the *APOE* region may influence neuropathology risk independently of the effects of *APOE* $\epsilon$ alleles.

To test the hypothesis that genetic loci in the *APOE* region may affect neuropathology risk independently from *APOE* $\epsilon$ alleles, we re-analyzed variants in Chromosome 19 while adjusting for *APOE* $\epsilon$ genotype. We limited re-analysis to endophenotypes with at least one significant association signal within the *APOE* locus in the pooled single-variant association analysis described in the previous section. *APOE* genotypes were determined using the rs7412 and rs429358 variants according to the SNPedia online reference [@apoe-s]. The $\epsilon3/\epsilon3$ genotype was used as reference, and we included fixed-effect indicator variables to adjust for $\epsilon2/\epsilon2$, $\epsilon2/\epsilon3$, $\epsilon3/\epsilon4$, and $\epsilon4/\epsilon4$ genotypes. We chose this approach rather than adjusting for counts of $\epsilon 2$ and $\epsilon 4$ alleles because it is robust to potential non-linear effects of genotypes.

### Gene-based analyses

Analyses that test for the aggregate effect of all variants in a gene region may increase power for detection of risk genes. We thus performed gene-based analyses for all neuropathology endophenotypes studied. Gene start and end sites were determined using GRCh38 gene regions. Variants were mapped to a gene if they were located within 10 kbp of the gene's start or end sites. Using an unrelated set of the pooled participant data set, gene-based analyses were performed using MAGMA v1.10 [@leeuw2015]. The same fixed-effects covariates used in single-variant analyses were used, and the top variant PCs that accounted for 99.9% of variance in a gene's region were used to test for significance using an F-test. We considered genes with $p < 2.5 \times 10^{-6}$ to be significantly associated with endophenotypes.

### Principal component outcome analyses

Many neuropathology endophenotypes are highly correlated with one another, and this phenomenon is not reflected in analyses using individual neuropathology endophenotypes as outcomes. Thus, we sought to perform genetic association analyses that could better account for the common co-occurrence of NPE. We first assessed the phenotype correlation of neuropathology endophenotypes used in the pooled GWAS using polychoric correlation and grouped endophenotypes by visually examining a dendrogram based on phenotype-phenotype correlations. Using the psych R package [@revelle2022], for each group of endophenotypes, we then performed PCA on the matrix of neuropathology endophenotype outcomes in that group. The first PC (PC1) of each group was then used as the outcome variable for GWAS using SAIGE in a two-stage analysis. In the first stage, PC1 was regressed on fixed-effects covariates, with the GRM included as a random effect. Because the residuals of PC1 may not be normally distributed, the residuals of stage one were then rank inverse normalized and regressed on covariates and GRM again in stage two. Score tests were then performed on genetic variants and a saddle-point approximation used to estimate p-values.

@@

Colocalization analysis is a Bayesian analytical method that seeks to answer whether a single genetic locus drives genotypic-phenotypic associations in two or more traits. We used multiple sources of publicly available summary statistics from external studies as data sources for colocalization analyses. First, we downloaded Genotype-Tissue Expression Project (GTEx) v8 European ancestry quantitative trait loci (QTL) analysis summary statistics, which contains summary statistics for significant gene expression and splicing QTL variants (eQTL and sQTL, respectively) in 48 body tissues [@gtexconsortium2017]. As an additional source of QTL data, we used gene expression and DNA methylation QTL (mQTL) analysis summary statistics from studies using tissue taken from the dorsolateral prefrontal cortex of ROSMAP participants [@ng2017]. These studies examined the associations of genetic variants with molecular traits and provide curated lists of significant QTL variants. Finally, we downloaded the summary statistics from a recent GWAS of LOAD for a targeted *post hoc* colocalization analysis in the *TMEM106B* and *GRN* genes [@bellenguez2022].

For each neuropathology endophenotype outcome in our study, we first compiled a list of genetic variants with p-values $\le 1 \times 10^{-5}$ in the pooled single-variant genome-wide association analysis. We then queried the lists of significant QTL variants in GTEx and ROSMAP to identify neuropathology-associated QTL variants. For each genetic locus associated with neuropathology endophenotypes that had at least one significant QTL in either GTEx or ROSMAP, we performed colocalization analysis using the "coloc.abf" function in the coloc R package [@giambartolomei2014]. For ordinal variables, we chose dichotomizing cut points to determine case-control proportions. We used coloc's default prior probability of colocalization (PrC) of $\text{PrC} = 1\times 10^{-5}$ and considered a posterior $\text{PrC}> 80\%$ as a threshold for evidence of colocalization.

To investigate whether shared GWAS signals drive association among multiple neuropathology endophenotypes, we also performed colocalization analysis on loci with variants exceeding a p-value threshold of $p < 1 \times 10^{-4}$ and concordant effect direction for at least two NPE in the pooled single-variant analysis.

# Results

```{r}
covar <- fread("data/mega/mega_np.covar")
nmega <- formatC(covar[, .N], big.mark = ",")
nnacc <- formatC(covar[FID == "0", .N], big.mark = ",")
nrosmap <- formatC(covar[FID == "1", .N], big.mark = ",")
nact <- formatC(covar[FID == "2", .N], big.mark = ",")
nadni <- formatC(covar[FID == "ADNI", .N], big.mark = ",")
```

```{r}
t1_env <- new.env()
source("code/manuscript/01_table1_participant_demographics.R",
       local = t1_env)
t1dt <- t1_env$t1dt
```

```{r}
t1_env$t1
```

In total, `r nmega` unique participants across all cohorts met our inclusion criteria and were included in our analyses; specifically, autopsy, and genotype data were available for these participants, and they passed our quality control measures for variant missingness, genetic heterogeneity, and ancestry. Included participants consisted of `r nnacc` participants from NACC; `r nrosmap` from ROSMAP; `r nact` from ACT; and `r nadni` from ADNI. Participant demographics are shown in **Table 1**. NACC participants died at a younger age (mean age at death 81 years) compared to ROSMAP (90 years) and ACT (88 years) participants, and had more balanced sex ratios, with 50% of NACC participants female vs. 68% and 58% in ROSMAP and ACT, respectively. NACC participants were also more likely to have an *APOE* $\epsilon 4$ allele (55%) vs. ROSMAP (25%) or ACT (28%). Across the 11 NPE harmonized across multiple cohorts, NACC and ADNI tended to show more advanced neuropathology than ROSMAP and ACT. Using CERAD neuritic plaque score as an example, 65% of NACC and 59% of ADNI participants were rated as "none" for CERAD score, whereas 33% of ROSMAP and 27% of ACT participants were rated as "severe." This trend can be seen in most of the NPE harmonized across cohorts with the exceptions of CAA, LATE-NC, and microinfarcts (see **Supplementary Table S2** for summary demographics for all neuropathology endophenotypes).

## Single-variant analysis identifies novel neuropathology-associated mapped to *COL4A1* and *PIK3R5*

```{r}
t2_env <- new.env()
source("code/manuscript/02_table2_mega_analysis_results.R",
       local = t2_env)
```

In the initial analysis using only NACC participants, variants within the *APOE* region were significantly associated with Braak NFT stage, CERAD score, CAA, and diffuse amyloid plaques. These associations were validated in all of the other data sets at a significance threshold of $p \le 0.05$. No other loci were significantly associated with any neuropathology endophenotypes.

In total, 14 independent ($r^2 < 0.05$) loci had lead variants that met the genome-wide significance threshold of $p < 5 \times 10^{-8}$ in the main pooled mega-analysis. Of these, four were duplicate signals in the *APOE* region, leaving 10 significant signals across 11 endophenotypes. Novel variant-endophenotype associations included (1) a variant 12 kbp upstream of *COL4A1,* rs2000660, associated with atherosclerosis in the circle of Willis (**Figure 1**) and (2) a *PIK3R5* intronic variant, rs72807981, associated with Braak NFT stage (**Figure 2**). Additionally, one locus with lead variant rs4803778 remained significantly associated with CAA in the *APOE* $\epsilon$-adjusted analysis (**Figure 3**). Loci identified that have been previously associated with NPE or dementia included *APOE* with Braak stage, CAA, CERAD score, diffuse amyloid plaques, and LATE-NC; *BIN1* with Braak NFT stage; and *TMEM106B* with HS and LATE-NC (**Table 2**).

```{r}
t2_env$t2
```

## Gene-based analysis confirms association between genes in *APOE* region and multiple neuropathology endophenotypes

Gene-level analyses further corroborated our single-variant analyses, identifying gene associations with neuropathology endophenotypes in regions with significant variant-phenotype associations. *APOE* was significantly associated with NFT, diffuse plaques, CAA, neuritic plaques, and LATE-NC. *CYP27C1*, located approximately 50 kbp downstream of *BIN1*, was significantly associated with NFT. *TMEM106B* was associated with both HS and LATE-NC.

## Analysis of neuropathology endophenotypic correlations identifies Alzheimer's-associated and vascular clusters
